Using Incremental Many-to-One Queries to Build a Fast and Tight Heuristic for A* in Road Networks

### 2021-11-10 paper/ch_potentials.tex
rewrite bidir dijkstra section

We consider directed graphs $G=(V,E)$ with $n=|V|$ nodes and $m=|E|$ edges with weight functions $w : E \to \mathbb{R}$.
Given nodes $s$ and $t$ we want to obtain a path $P=(s=v_0,\dots,t=v_k)$ of minimal weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
We denote this shortest path weight as the distance $\dist_w(s,t)$ between $s$ and $t$.

@@

@@

Dijkstra's algorithm can also be run from the target $t$ on the reversed graph.
We call this a \emph{backward search}.
Running two Dijkstra searches, one from $s$ and a backward search from $t$, until the searches meet is called a \emph{Bidirectional Dijkstra}.
Typically, the searches are interleaved by alternating settling a node from each direction.
Another common approach is to always settle a node from the direction with the smaller minimal queue key.
Theoretically any direction interleaving strategy is possible.
The searches can also be run in parallel in two threads.
When the sum of the minimum keys in both queues $\overrightarrow{k} + \overleftarrow{k}$ is greater than the so far best known tentative total distance $\mu$ the algorithm can terminate.

@@

Since bidirectional search provides an easy way to halve the running time of Dijkstra's algorithm, a bidirectional variant of A* also seems desirable.
However, as pointed out in~\cite{gh-cspas-05}, this is not as straightforward as one might expect.
One would like to use a forward potential $\overrightarrow{h}_t$ in the forward search and a backward potential $\overleftarrow{h}_s$ (computing distances in the reversed graph) in the reverse search.
The problem is that these two potentials induce different reduced graphs (see Section~\ref{sec:a_star}).
Thus, in the equivalent bidirectional Dijkstra search each direction would run on a different graph.
This breaks the bidirectional Dijkstra stopping criterion.
To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until their minimum queue key $k$ becomes greater than the best known total tentative distance $\mu$.
But this is the same stopping criterion that a unidirectional A* search has.
Thus, this straightforward bidirectional A* performs twice the work than a unidirectional A*.
This can be slightly improved by introducing pruning.
Nodes for which $\overrightarrow{D}[x] + h(x) > \mu$ holds, i.e. their distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, can not contribute to a shorter path and need not to be added to the queue.
This approach is denoted as the \emph{symmetric} bidirectional A*.

To overcome this problem the authors of~\cite{gh-cspas-05} propose the \emph{average} potential, a combined heuristic, which is good for both directions but also makes both searches run on the same reduced graph and thus enables an efficient stopping criterion.
The reduced graph of the average potential is the graph with the average weights of the both reduced graphs.
The potential function which induces this average reduced graph is $h_{\overline{st}}(x) := \frac{\overrightarrow{h}_t(x) - \overleftarrow{h}_s(x)}{2}$ (and $-h_{\overline{st}}(x)$ for the backward search).
This potential allows stopping the searches when $\overrightarrow{k} + \overleftarrow{k} \geq \mu$.
Additionally, the pruning from the symmetric approach can be utilized with the original potentials.
To the best of our knowledge, this method is the fastest known variant of bidirectional A*.
Still, it has two downsides.
First, evaluating the average potential requires evaluating the original heuristics for both directions.
This can cause some overhead depending on the heuristics.
Second, for each direction on its own, the average potential is a worse heuristic than the original heuristic.

We propose an improved pruning criterion for the symmetric approach.
It takes the search progress (in form of the growing minimal queue keys) into account to tighten the pruning with increasing search progress and compensates for the worse stopping criterion.
As our experiments show, this makes the simpler symmetric approach competitive to the average potential.
Recall that the queue is ordered by $D[x] + h(x)$.
We use the minimal queue key $\overleftarrow{k}$ of the backward search to obtain a tighter estimate of $\dist(x,t)$.
If a node $x$ was not yet settled by backward search then $\dist(x,t) \geq \overleftarrow{k} - \overleftarrow{h}_s(x)$.
Thus, in the forward search if $\overrightarrow{D}[x] + \overleftarrow{k} - \overleftarrow{h}_s(x) > \mu$, we do not need to push $x$ into the queue.
The pruning for the backwards search works analogously.
Since this pruning is only used when $\mu > \infty$, for the greatest part of the search each direction only has to evaluate its own heuristic.

For both average and symmetric bidirectional A*, we always alternate between settling a node form the forward and the backward search.
In our experiments, we also evaluate choosing the direction by the smaller minimal queue key.
Even though this appears to better fit together with the stopping criterion for the symmetric approach, our experiments show that it is never beneficial in terms of running time (see Table~\ref{tab:bidir_switching}).

@@

Table~\ref{tab:bidir_switching} shows the performance of bidirectional A* with different strategies to decide whether to advance the forward or the backward search next.
The results clearly show that alternating the directions is always superior.
Selecting the direction by minimal queue key may lead to huge imbalances in the progress of the searches.
This causes the searches to meet later and the total search space to grow significantly.

>>>>

We consider directed graphs $G=(V,E)$ with $n=|V|$ nodes and $m=|E|$ edges with weight functions $w : E \to \mathbb{R}$.
Given nodes $s$ and $t$ we want to obtain a path $P=(s=v_0,\dots,t=v_k)$ of minimum weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
We denote this shortest path weight as the distance $\dist_w(s,t)$ between $s$ and $t$.

@@

% TODO Capital letters for arrays?

@@

Dijkstra's algorithm can also be run from the target $t$ on the reversed graph.
We call this a \emph{backward search}.
Running two Dijkstra searches, one from $s$ and a backward search from $t$, until the searches meet is called a \emph{Bidirectional Dijkstra}.
Typically, the searches are interleaved by alternating settling a node from each direction.
Another common approach is to always settle a node from the direction with the smaller minimum queue key.
Theoretically any direction interleaving strategy is possible.
The searches can also be run in parallel in two threads.
When the sum of the minimum keys in both queues $\overrightarrow{k} + \overleftarrow{k}$ is greater than the so far best known tentative total distance $\mu$ the algorithm can terminate.

@@

On road graphs, bidirectional search provides an easy way to halve the running time of Dijkstra's algorithm.
A bidirectional variant of A* therefore appears also desirable.
However, as shown in~\cite{gh-cspas-05}, this is not as simple as it seems.

The straightforward idea would be to use two heuristics.
The forward search has its queue ordered by $\overrightarrow{D}[x] + \overrightarrow{h}_t(x)$ where $\overrightarrow{h}_t(x)$ estimates the distance from $x$ to $t$.
Similarly, the backward search has its queue ordered by $\overleftarrow{D}[x] + \overleftarrow{h}_s(x)$ where $\overleftarrow{h}_s(x)$ is an estimate of the distance from $s$ to $x$.
% We denote by $\overrightarrow{D}[x]$ the tentative distance of the forward search at $x$ and similarly, by $\overleftarrow{D}[x]$ the tentative distance in the backward search.
We denote by $\overrightarrow{k}$ and $\overleftarrow{k}$ the minimum queue keys and by $\mu$ the length of the shortest $st$-path found so far.

The problem with this straightforward approach is that these two potentials induce different reduced graphs (see Section~\ref{sec:a_star}).
Thus, in the equivalent bidirectional Dijkstra search each direction would run on a different graph.
This breaks the usual bidirectional Dijkstra stopping criterion.
To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until the unidirectional stopping criterion is met for one direction.
Thus, this straightforward bidirectional A* performs roughly twice the work of a unidirectional A*.
The authors of~\cite{gh-cspas-05} refer to this approach as \emph{symmetric} bidirectional A*.

To obtain a better stopping criterion, the \emph{average potential} is proposed~\cite{gh-cspas-05}.
It combines a forward and a backward heuristic $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$ into a combined heuristic $h_{\overline{st}}$, which is good for both directions.
Both searches run on the same reduced graph which enables an efficient stopping criterion.
The reduced graph of the average potential is the graph with the average weights of the reduced graphs of $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$.
The heuristic function which induces this average reduced graph is $h_{\overline{st}}(x) := \frac{\overrightarrow{h}_t(x) - \overleftarrow{h}_s(x)}{2}$.
The backward search uses $-h_{\overline{st}}(x)$.
This allows stopping the searches when $\overrightarrow{k} + \overleftarrow{k} \geq \mu$.
To the best of our knowledge, this method is the fastest known variant of bidirectional A*.
Unfortunately, it has two downsides.
First, evaluating the average potential requires evaluating both $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$.
This makes it more expensive in terms of running time.
% Second, for each direction on its own, the average potential is a worse heuristic than the original heuristic.
Second, the better stopping criterion comes at the cost of worse estimates for each direction on its own.
The average potential estimates $\dist(x,t)$ less accurately than $\overrightarrow{h}_t(x)$ and $\dist(s,x)$ less accurately than $\overleftarrow{h}_s(x)$.
The second downside can be partially mitigated through pruning with the original heuristics.
When the forward search scans an edge $xy$ where $\overrightarrow{D}[x] + w(xy) + \overrightarrow{h}_s(y) > \mu$ holds, i.e. the distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, $y$ needs not to be inserted into the queue.
The pruning for the backward search works analogously.

To avoid the downsides of the average potential, we revisit symmetric bidirectional A* and propose a new pruning criterion.
We use the minimum queue key $\overleftarrow{k}$ of the backward search to obtain a tighter estimate of $\dist(x,t)$ for pruning in the forward search.
Recall that queues are ordered by $D[x] + h(x)$.
If a node $x$ was not yet settled by backward search then $\dist(x,t) + \overleftarrow{h}_s(x) \geq \overleftarrow{k}$, and equivalently $\dist(x,t) \geq \overleftarrow{k} - \overleftarrow{h}_s(x)$.
Thus, in the forward search if $\overrightarrow{D}[x] + \overleftarrow{k} - \overleftarrow{h}_s(x) > \mu$, we do not need to push $x$ into the queue.
The pruning for the backwards search works analogously.
As the pruning becomes tighter as the searches progress, it compensates for the worse stopping criterion.
Since this pruning is only used when $\mu > \infty$, for the greatest part of the search each direction only has to evaluate its own heuristic.
As our experiments show, this makes the simpler symmetric approach competitive to the average potential.

Unless stated differently, for all bidirectional A* variants, we always alternate between removing a node from the forward and the backward queues.
In our experiments, we also evaluate expanding the search with the smaller minimum queue key.
This sounds sensible but our experiments show in Table~\ref{tab:bidir_switching} that it is never beneficial in terms of running time.

@@

Table~\ref{tab:bidir_switching} shows the performance of bidirectional A* with different strategies to decide whether to advance the forward or the backward search next.
The results clearly show that alternating the directions is always superior.
Selecting the direction by minimum queue key may lead to huge imbalances in the progress of the searches.
This causes the searches to meet later and the total search space to grow significantly.
